---
title: "CS 598 Practical Statistical Learning Coding Assignment #3"
output: html_document
---
### Team Members
#####  Derek Chapman (derek4)
#####  Zeed Jarrah (zjarrah2)  
<br>

####  Helper Functions

1) lo.lev
The smoother matrix is computed and used to store the diagonal values in the vector lev.
The method for recreating the smoother matrix uses the identity matrix Id_mat and
performing matrix multiplication on the matrix containing the predicted values as seen using the method predict(loess()).

2) onestep_CV
The residual error is computed separately and stored as a vector to be used in the computation 
of LOO-CV and GCV. We note that the distinguishing factor between the two are that we use mean
leverage when computing GCV.

3) myCV
This function calls onestep_CV and is used to compute LOOCV and GCV for every value of span in the sequence.

```{r prep_func}
lo.lev <- function(x1, sp){
  # x1: feature vector of length n
  # sp: a value for "span"
  
  n = length(x1)
  
  S_mat = matrix(0,n,n)
  Id_mat = diag(n)
  
  #Fitted values
  for(i in 1:n){
    S_mat[,i] = predict(loess(Id_mat[,i] ~ x1, span = sp))
  }
  
  #leverage values are diagonals of S-lambda
  lev = diag(S_mat)
  
  return(lev)
}

onestep_CV <- function(x1, y1, sp){
  fit1 = loess(y1 ~ x1, span = sp)
  resid_vec = resid(fit1)
  S_vec = lo.lev(x1, sp)
  
  # LOO-CV
  cv = mean((resid_vec/(1-S_vec))^2)
  
  # GCV
  mean_lev = mean(S_vec)
  gcv = mean((resid_vec/(1-mean_lev))^2)
  
  return(list(cv = cv, gcv = gcv))
}

myCV <- function(x1, y1, span){
  # x1: feature vector of length n
  # y1: response vector of length n
  # span: a sequence of values for "span"
  
  m = length(span)
  cv = rep(0, m)
  gcv = rep(0, m)
  
  for(i in 1:m){
    tmp = onestep_CV(x1, y1, span[i])
    cv[i] = tmp$cv
    gcv[i] = tmp$gcv
  }
  return(list(cv = cv, gcv = gcv))
}
```
<em>
Plotting the fitted curve against the observed data values.
<em/>
```{r fig.align="center"}
# Testing Function
mydata = read.csv(file = "Coding3_Data.csv")
dim(mydata)

plot(mydata$x, mydata$y, xlab="", ylab="")

span1 = seq(from = 0.2, by = 0.05, length = 15 )
cv.out = myCV(mydata$x, mydata$y, span1)

# Print Results
myout = data.frame(CV = cv.out$cv, 
                   GCV = cv.out$gcv, 
                   span = span1)
myout

myout$span[myout$GCV == min(myout$GCV)]

myout$span[myout$CV == min(myout$CV)]
```

```{r fig.align="center"}
# Plot The Fitted Curve
spangcv.min = myout$span[myout$GCV == min(myout$GCV)]
spancv.min = myout$span[myout$CV == min(myout$CV)]

plot(mydata$x, mydata$y, 
     xlab="", ylab="", 
     main = "True curve and Local Regression fitted curve with Span of 0.5",
     col="gray")
fx = 1:50/50;
fy = sin(12*(fx+0.2))/(fx+0.2)
lines(fx, fy, col=8, lwd=2)

f = loess(y ~ x, mydata, span = spangcv.min)
lines(fx, predict(f, data.frame(x = fx), surface = "direct"), 
      lty=2, lwd=2, col="blue")
```
